\texttt{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-08-24 dfoliatR_ms.Rmd
Last edits before resubmission/acceptance

The `dfoliatR` package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes.  It is built on the long-accepted host to non-host comparison methodology used in the 1980s FORTRAN program OUTBREAK [@Swetnam1985; @Swetnam1989; @Swetnam1993]. Our evaluation of the two programs revealed that `dfoliatR` excelled in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. `dfoliatR` provides easier control of the rule base for suppression thresholds, additional output tables, and high-quality and customizable graphics. These features allow users to compare regimes of different tree species or geographic regions, evaluate sample-size considerations, examine a multitude of relevant insect disturbance questions, and more readily test the effects of different rule-base thresholds on the likelihood of Type I and II errors. Finally, `dfoliatR` operates in the open-source `R` environment that is stable across computing platforms and is under active development and maintenance by a large and growing community.

>>>>

The `dfoliatR` package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes.  It is built on the long-accepted host to non-host comparison methodology used in the 1980s FORTRAN program OUTBREAK [@Swetnam1985; @Swetnam1989; @Swetnam1993]. Our evaluation of the two programs revealed that `dfoliatR` excels in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. `dfoliatR` provides easier control of the rule base for suppression thresholds, additional output tables, and high-quality and customizable graphics. These features allow users to compare insect outbreak regimes of different tree species or geographic regions, evaluate sample-size considerations, examine a multitude of relevant insect disturbance questions, and more readily evaluate the potential for Type I and II errors in their results. Finally, `dfoliatR` operates in the open-source `R` environment that is stable across computing platforms and is under active development and maintenance by a large and growing community.

### 2020-08-24 dfoliatR_ms.tex
Last edits before resubmission/acceptance

The main dendrochronological tool for inferring, dating, and characterizing defoliator outbreaks from tree-ring records has been the software routine OUTBREAK (Swetnam et al., 1985; Holmes and Swetnam, 1986; Swetnam and Lynch, 1989). OUTBREAK computes indices (described later in detail) of suppressed growth by subtracting a detrended and standardized climate series (a ``control'' chronology) from individual host-tree detrended and standardized radial growth series after the host and non-host series have been brought to a common variance. The non-host chronology usually consists of a site chronology developed from non-host tree species growing on a climate-sensitive site, but a gridded climate data point series, like the North American Drought Atlas (Cook and Krusic, 2004) also suffices. If the host and non-host species respond similarly to climate (which can and should be tested), the derived series retains variability that the host and non-host series do not have in common, generally the insect signal and some unexplained variability (noise). The user defines a rule base specifying the magnitude and duration that a period of indexed growth suppression must meet or surpass for a period of suppressed growth to be inferred as a defoliation event at the tree level.

@@

The \texttt{dfoliatR} package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes. It is built on the long-accepted host to non-host comparison methodology used in the 1980s FORTRAN program OUTBREAK (Swetnam et al., 1985; Swetnam and Lynch, 1989, 1993). Our evaluation of the two programs revealed that \texttt{dfoliatR} excelled in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. Other key benefits are that \texttt{dfoliatR} provides easier control of suppression thresholds, additional output tables, high-quality and customizable graphics, and operates in the open-source \texttt{R} environment that is stable across computing platforms and is under active development and maintenance by a large and growing community.

@@

\leavevmode\hypertarget{ref-dfoliatR}{}%
Guiterman, C., Lynch, A., Axelson, J., 2020. \texttt{dfoliatR}: Detection and Analysis of Insect Defoliation Signals in Tree Rings., \url{https://chguiterman.github.io/dfoliatR/}, doi:\href{https://doi.org/10.5281/zenodo.3698163}{10.5281/zenodo.3698163}.

>>>>

The main dendrochronological tool for inferring, dating, and characterizing defoliator outbreaks from tree-ring records has been the software routine OUTBREAK (Swetnam et al., 1985; Holmes and Swetnam, 1986; Swetnam and Lynch, 1989). OUTBREAK computes indices (described later in detail) of suppressed growth by subtracting a detrended and standardized climate series (a ``control'' chronology) from individual host-tree detrended and standardized radial growth series after the host and non-host series have been brought to a common variance. The non-host chronology usually consists of a site chronology developed from non-host tree species growing on a climate-sensitive site, but a gridded climate data point series, like the North American Drought Atlas (Cook and Krusic, 2004) also suffices. If the host and non-host species respond similarly to climate (which can and should be tested), the derived series retains variability that the host and non-host series do not have in common, generally the insect signal and some unexplained variability (noise). The user defines a rule base specifying the magnitude and duration that a period of indexed growth suppression must meet or surpass for a period of suppressed growth to be inferred as a defoliation event at the tree level. Rule bases are derived from the user's knowledge of insect and host ecologies, and from consideration of the likelihood and relative importance of Type I and II errors.

@@

The \texttt{dfoliatR} package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes. It is built on the long-accepted host to non-host comparison methodology used in the 1980s FORTRAN program OUTBREAK (Swetnam et al., 1985; Swetnam and Lynch, 1989, 1993). Our evaluation of the two programs revealed that \texttt{dfoliatR} excels in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. \texttt{dfoliatR} provides easier control of the rule base for suppression thresholds, additional output tables, and high-quality and customizable graphics. These features allow users to compare insect outbreak regimes of different tree species or geographic regions, evaluate sample-size considerations, examine a multitude of relevant insect disturbance questions, and more readily evaluate the potential for Type I and II errors in their results. Finally, \texttt{dfoliatR} operates in the open-source \texttt{R} environment that is stable across computing platforms and is under active development and maintenance by a large and growing community.

@@

\leavevmode\hypertarget{ref-dfoliatR}{}%
Guiterman, C., Lynch, A., Axelson, J., 2020. \texttt{dfoliatR}: Detection and Analysis of Insect Defoliation Signals in Tree Rings. R package version 0.2.0., \url{https://chguiterman.github.io/dfoliatR/}, doi:\href{https://doi.org/10.5281/zenodo.3626135}{10.5281/zenodo.3626135}.
